---
title: "Project 2- Data Tidying"
author: "Kiera Griffiths, Desiree Thomas"
format: html
editor: source
---

# Approach 
For Project 2, we will be tidying 3 datasets. This file focuses on the Generational Takeout Spending dataset. We will import the wide-format CSV file from GitHub into R and transform it into a tidy structure using the tidyr and dplyr packages. We will reshape the generational spending columns into a long format using functions such as pivot_longer(), then standardize column names with rename_with() and handle any missing or inconsistent values using functions like drop_na() to prepare the data for analysis. Finally, we will analyze trends in takeout spending across generations and visualize the results using ggplot2, though a potential challenge may be restructuring the generational columns correctly during the wide-to-long transformation.

```{r}
library(tidyr)
library(dplyr)
```

# Data Source
We created the dataset based on studies showing the takeout spending habits across various generations. Imported this raw CSV file from the main branch of our project 2 public GitHub repository. 
```{r}
generational_takeout <- read.csv("https://raw.githubusercontent.com/meiqing39/607-Project-2/refs/heads/main/Generational%20Takeout%20Spending.csv")

```

# Data Structure Before Tidying
To perform an initial inspection of the dataset, we printed the generational_takeout tibble to the console. This provided a summary of the data structure, confirming a dimension of 5 rows and 7 columns. By viewing the data in this format, we were able to verify that the spending values were correctly imported and that the dataset currently exists in a wide format.
```{r}
generational_takeout
```


```{r}

# Reshape to tidy format
tidy_gen_takeout <- generational_takeout |>

  pivot_longer(

    cols      = -id,

    names_to  = "variable",

    values_to = "value"

  ) |>

  rename_with(tolower) |>

  drop_na()

```

