---
title: "Project 2- Data Tidying"
author: "Kiera Griffiths, Desiree Thomas"
format: html
editor: source
---

# Approach 
For Project 2, we will be tidying 3 datasets. This file focuses on the Generational Takeout Spending dataset. We will import the wide-format CSV file from GitHub into R and transform it into a tidy structure using the tidyr and dplyr packages. We will reshape the generational spending columns into a long format using functions such as pivot_longer(), then standardize column names with rename_with() and handle any missing or inconsistent values using functions like drop_na() to prepare the data for analysis. Finally, we will analyze trends in takeout spending across generations and visualize the results using ggplot2, though a potential challenge may be restructuring the generational columns correctly during the wide-to-long transformation.

```{r}
library(tidyr)
library(dplyr)
library(ggplot2)
```

# Data Source
We created the dataset based on studies showing the takeout spending habits across various generations. Imported this raw CSV file from the main branch of our project 2 public GitHub repository. 
```{r}
generational_takeout <- read.csv("https://raw.githubusercontent.com/meiqing39/607-Project-2/refs/heads/main/Generational%20Takeout%20Spending.csv")

```

# Data Structure Before Tidying
To perform an initial inspection of the dataset, we printed the generational_takeout tibble to the console. This provided a summary of the data structure, confirming a dimension of 5 rows and 7 columns. By viewing the data in this format, we were able to verify that the spending values were correctly imported and that the dataset currently exists in a wide format.
```{r}
generational_takeout
```
# Transformation
We reshaped the dataset from wide to tidy (long) format. We normalized variable structures and renamed variables to follow a consistent naming convention. We did not have missing or inconsistent values to address.
```{r}

tidy_gen_takeout <- generational_takeout %>%
  
  # Step 1: Reshape from wide to long (tidy) format
  pivot_longer(
    cols = -Generation,       # all columns except 'Generation'
    names_to = "period",      # column names (Jan.25, Feb.25...) go into 'period'
    values_to = "expenses"    # values go into 'expenses'
  ) %>%
 
  # Step 2: Split 'period' into month and year, this is part of the normalization process.
  #\\. is used to escape the '.' and prevent an error in the separate() function; '.' on it's own is a wildcard character 
  separate(period, into = c("month", "year"), sep = "\\.") %>%
 
   # Step 3: Standardize year to 4 digits and convert Generation to factor
  mutate(
    year = paste0("20", year),
    Generation = as.factor(Generation)
  ) %>%
 
   # Step 4: Standardize column names manually
  rename(
    generation = Generation
  )

# Preview tidy dataset
View(tidy_gen_takeout)

```

# Analysis
We provided two analyses for our dataset. First we wanted to calculate each generation’s percentage of each month's total spending. We loaded tidy_gen_takout into the dataframe gen_monthly_percentage, calculated the percentages then plotted the results in a bar graph and summary table. 

The next analysis calculated the average monthly spending for each generation, over the entire 6-month period. We loaded the tidy_gen_takeout into a dataframe named generation_analysis. We ranked the generations in descending order and plotted the results in a bar graph and summary table for visualization.
```{r}
# Generational Share of Monthly Takeout Spending

gen_monthly_percentage <- tidy_gen_takeout %>%
  group_by(month) %>%                       # group by month
  mutate(total_month = sum(expenses)) %>%  # total spending across all generations per month
  ungroup() %>%
  mutate(percent = round((expenses / total_month) * 100)) %>%  # round to whole %
  select(generation, month, expenses, percent)

gen_monthly_percentage
```
```{r}
# Month order
month_levels <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun")

# Convert month column to ordered factor
gen_monthly_percentage <- gen_monthly_percentage %>%
  mutate(month = factor(month, levels = month_levels))

ggplot(gen_monthly_percentage, aes(x = month, y = percent, fill = generation)) +
  geom_bar(stat = "identity", position = "stack") +   # stacked by generation
  geom_text(aes(label = paste0(percent, "%")),
            position = position_stack(vjust = 0.5),   # labels inside each bar
            size = 3) +
  labs(
    title = "Generational Percentage of Monthly Takeout Spending",
    x = "Month",
    y = "Percentage of Total Monthly Spending (%)",
    fill = "Generation"
  ) +
  theme_minimal()
```

```{r}
library(gt)

# 1️⃣ Prepare summary table: months as rows, generations as columns
summary_table <- gen_monthly_percentage %>%
  select(month, generation, percent) %>%
  pivot_wider(names_from = generation, values_from = percent) %>%
  arrange(match(month, c("Jan","Feb","Mar","Apr","May","Jun")))

# 2️⃣ Create styled summary chart with gt
summary_table %>%
  gt() %>%
  tab_header(
    title = "Generational Percentages per Month"
  ) %>%
  # 3️⃣ Add alternating row colors
  tab_style(
    style = list(
      cell_fill(color = "lightgray")
    ),
    locations = cells_body(
      rows = seq(1, nrow(summary_table), 2)  # every other row
    )
  ) %>%
  # 4️⃣ Optional: center-align text
  cols_align(
    align = "center",
    columns = everything()
  )
```

```{r}
generational_analysis <- tidy_gen_takeout %>%

# 3. Target: Group by Generation
  group_by(generation) %>%

# 4. Focus: Calculate the Average spending across the months
  summarise(avg_monthly_spending = round(mean(expenses))) %>%
  
# 5. Rank: Order from highest to lowest
  arrange(desc(avg_monthly_spending))
 
# View the final ranked results
print(generational_analysis)
```

```{r}

library(ggplot2)

ggplot(generational_analysis, 
       aes(x = reorder(generation, avg_monthly_spending),
           y = avg_monthly_spending, 
           fill = generation)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0("$", avg_monthly_spending)),
            vjust = -0.5, size = 4) +
  labs(
    title = "Average Monthly Takeout Spending by Generation",
    x = "Generation",
    y = "Average Monthly Spending ($)",
    fill = "Generation"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```

```{r}
library(gt)

generational_analysis %>%
  gt() %>%
  tab_header(
    title = "Average Monthly Takeout Spending by Generation"
  ) %>%
  cols_align(
    align = "center",
    columns = everything()
  ) %>%
  tab_style(
    style = cell_fill(color = "lightgray"),
    locations = cells_body(
      rows = seq(1, nrow(generational_analysis), 2)
    )
  )
```

